---
title: "U.S. Car Accidents (2016-2023)"
output:
  html_document:
    df_print: paged
---

```{r, global_options, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, tidy.opts = list(width.cutoff = 60))
suppressPackageStartupMessages(library(ggplot2))
```

```{r, echo=FALSE}
library(readr)
library(here)
library(dplyr)
library(stringr)
library(lubridate)
library(ggplot2)
library(plotly)
library(ggiraphExtra)
library(maps)
library(mapproj)
library(xts)  
library(dygraphs)
library(reshape2)
library(kableExtra)


# -- Import and Clean Data --
if (!exists("accidents_raw")) {
  accidents_raw <- read_csv(here("data", "raw_data", "US_Accidents_March23.csv"),
    lazy = FALSE
  )
}

acc <- accidents_raw

# cleanup
acc <- acc %>%
  filter(between(`Temperature(F)`, -60, 130)) %>%
  mutate(
    date_ = date(Start_Time),
    year_ = year(Start_Time),
    month_ = month(Start_Time),
    hour_ = hour(Start_Time),
    precipitation = coalesce(`Precipitation(in)`, 0),
    any_precip = precipitation > 0,
    weather = coalesce(Weather_Condition, "Unknown"),
    temperature = coalesce(`Temperature(F)`, mean(`Temperature(F)`, na.rm = TRUE)),
    visibility = `Visibility(mi)`,
    wind_chill = `Wind_Chill(F)`
  )

# augmentation
state_lookup <- data.frame(
  State = state.abb,
  state_name = str_to_title(state.name)
)

# Create day_of_week column from date_
acc$day_of_week <- weekdays(acc$date_)

acc$day_of_week <- factor(acc$day_of_week,
                                levels = c("Sunday", "Monday", "Tuesday", "Wednesday", 
                                           "Thursday", "Friday", "Saturday"))

acc <- acc %>%
  mutate(
    sevg = case_when(
      Severity == 1 ~ "least severe",
      Severity == 2 ~ "less severe",
      Severity == 3 ~ "more severe",
      TRUE ~ "most severe"
    )
  ) %>%
  inner_join(state_lookup, by = "State")
```

# Introduction

# Descriptive Analysis

```{r, echo=FALSE}
# summary(acc)
# dim(acc)
# str(acc)
# head(acc)
```

The raw dataset contains `r scales::comma(nrow(accidents_raw))` observations (rows) of `r ncol(accidents_raw)` variables (columns).

After data preparation and cleaning, the dataset contains `r scales::comma(nrow(acc))` observations (rows) of `r ncol(acc)` variables (columns).

```{r, echo=FALSE}
# ACCIDENT SEVERITY COUNTS
sev_count <- acc %>%
  count(sevg)

p <- ggplot(sev_count, aes(x = sevg, y = n)) +
  geom_col() +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Accident Counts by Severity (2016 - 2023)",
    x = "Severity",
    y = "Number of Accidents"
  )

f <- here("Rmd", "ieee", "figures", "acc-by-severity.png")
ggsave(f, p +
         theme_minimal(base_size = 9), 
         width = 3.5, height = 2.5, units = "in")
knitr::plot_crop(f)

ggplotly(p)
knitr::kable(sev_count, col.names = c("Severity", "Number of Accidents"))
```

The author defines severity as "the impact on traffic." Low severity accidents would have a minimal effect on traffic whereas high severity accidents would have a significant impact on traffic.

We can observe that the majority of accidents that took place between 2016 and 2023 were categorized as "`r (sev_count %>% arrange(desc(n)) %>% head(1))$sevg`," accounting for `r scales::comma((sev_count %>% arrange(desc(n)) %>% head(1))$n)` of the total `r scales::comma(nrow(acc))` accidents.

## Time-Based Trends

### Daily Accident Trends (2016–2023)

The interactive time series plot shows daily accident counts across the United States from 2016 to 2023. The frequency of reported accidents increased noticeably after 2020, with peaks exceeding 10,000 accidents per day. This upward trend may reflect improved reporting mechanisms, changes in driving behavior, or broader shifts in traffic volume and weather conditions.

```{r}
accident_counts <- acc %>%
  group_by(date_) %>%
  summarise(count = n())

accident_xts <- xts(accident_counts$count, order.by = accident_counts$date_)

dygraph(accident_xts, main = "Daily Accident Counts (2016–2023)") %>%
  dySeries("V1", label = "Accidents") %>%
  dyRangeSelector()

```

### Average Accident Severity by Weather Condition

Among the top 10 most common weather conditions, "Overcast" and "Scattered Clouds" were associated with the highest average accident severity. In contrast, fair weather conditions such as "Fair" and "Fog" were linked to lower severity scores. This suggests that overcast or unstable weather may contribute to more serious traffic incidents.

```{r}
weather_severity <- acc %>%
  filter(!is.na(weather)) %>%
  group_by(weather) %>%
  summarise(
    avg_severity = mean(Severity),
    count = n()
  ) %>%
  arrange(desc(count)) %>%
  slice(1:10) 

# Bar plot
p <- ggplot(weather_severity, aes(x = reorder(weather, -avg_severity), y = avg_severity)) +
  geom_col(fill = "steelblue") +
  labs(title = "Average Severity by Weather Condition (Top 10)",
       x = "Weather Condition", y = "Average Severity") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

f <- here("Rmd", "ieee", "figures", "sev-by-weather.png")
ggsave(f, p +
  theme_minimal(base_size = 7) ,
         width = 3.5, height = 2.5, units = "in")
knitr::plot_crop(f)

ggplotly(p + theme_minimal())
```

### Accidents Distribtued by Hour, Day of Week, and Month

#### Hourly Distribution of Accidents

The distribution of accidents by hour reveals two major peaks: one around 7–8 AM and another between 3–6 PM, corresponding to typical rush hour periods. Fewer accidents occur during the early morning hours, while activity steadily increases throughout the day and decreases again in the evening.

```{r}
p <- ggplot(acc, aes(x = hour_)) +
  geom_bar(fill = "skyblue", color = "black") +
  labs(title = "Accidents by Hour of Day",
       x = "Hour (24-Hour Format)", y = "Number of Accidents") +
  scale_y_continuous(labels = scales::comma) 

f <- here("Rmd", "ieee", "figures", "acc-by-hour.png")
ggsave(f, p +
  theme_minimal(base_size = 7) ,
         width = 3.5, height = 2.5, units = "in")
knitr::plot_crop(f)

ggplotly(p + theme_minimal())
```

#### Weekly Distribution of Accidents

Accidents occurred most frequently on weekdays, with Friday showing the highest count, followed closely by Wednesday and Thursday. Sundays and Saturdays saw significantly fewer accidents. This pattern reflects increased commuting activity during the workweek compared to weekends.

```{r}
p <- ggplot(acc, aes(x = day_of_week)) +
  geom_bar(fill = "lightgreen", color = "black") +
  labs(title = "Accidents by Day of the Week",
       x = "Day", y = "Number of Accidents") +
  scale_y_continuous(labels = scales::comma)   # optional, for comma formatting

f <- here("Rmd", "ieee", "figures", "acc-by-weekday.png")
ggsave(f, p +
  theme_minimal(base_size = 7) ,
         width = 3.5, height = 2.5, units = "in")
knitr::plot_crop(f)

ggplotly(p + theme_minimal())
```

#### Monthly Distribution of Accidents

December experienced the highest number of accidents, followed by January and November. Accident frequency was generally lower in the summer months, particularly July. This trend may reflect seasonal variations such as holiday travel, winter weather conditions, or changes in daylight and visibility.

```{r}
# ----------- Accidents by Month -----------

p <- ggplot(acc, aes(x = factor(month_))) +
  geom_bar(fill = "orange", color = "black") +
  labs(title = "Accidents by Month",
       x = "Month", y = "Number of Accidents") +
  scale_y_continuous(labels = scales::comma) 

f <- here("Rmd", "ieee", "figures", "acc-by-month.png")
ggsave(f, p +
  theme_minimal(base_size = 7) ,
         width = 3.5, height = 2.5, units = "in")
knitr::plot_crop(f)

ggplotly(p + theme_minimal())
```

## Geographic Trends

```{r, echo=FALSE}
# https://www.census.gov/data/tables/time-series/demo/popest/2020s-state-detail.html
est_pop_state_2023 <- data.frame(
  state_name = c(
    "Alabama", "Alaska", "Arizona", "Arkansas", "California", "Colorado",
    "Connecticut", "Delaware", "District of Columbia", "Florida", "Georgia",
    "Hawaii", "Idaho", "Illinois", "Indiana", "Iowa", "Kansas", "Kentucky",
    "Louisiana", "Maine", "Maryland", "Massachusetts", "Michigan", "Minnesota",
    "Mississippi", "Missouri", "Montana", "Nebraska", "Nevada", "New Hampshire",
    "New Jersey", "New Mexico", "New York", "North Carolina", "North Dakota",
    "Ohio", "Oklahoma", "Oregon", "Pennsylvania", "Rhode Island",
    "South Carolina", "South Dakota", "Tennessee", "Texas", "Utah", "Vermont",
    "Virginia", "Washington", "West Virginia", "Wisconsin", "Wyoming"
  ),
  pop = c(
    5108468, 733406, 7431344, 3067732, 38965193, 5877610, 3617176, 1031890,
    678972, 22610726, 11029227, 1435138, 1964726, 12549689, 6862199, 3207004,
    2940546, 4526154, 4573749, 1395722, 6180253, 7001399, 10037261, 5737915,
    2939690, 6196156, 1132812, 1978379, 3194176, 1402054, 9290841, 2114371,
    19571216, 10835491, 783926, 11785935, 4053824, 4233358, 12961683, 1095962,
    5373555, 919318, 7126489, 30503301, 3417734, 647464, 8715698, 7812880,
    1770071, 5910955, 584057
  )
)

states_map <- map_data("state")

# make state names uppercase so they look better on the map
states_map$region <- str_to_title(states_map$region)

acc_state <- acc %>%
  group_by(State, state_name) %>%
  summarise(
    mean_sev = mean(Severity),
    n_acc = n(),
    mean_temp = mean(`Temperature(F)`),
    .groups = "drop"
  ) %>%
  left_join(est_pop_state_2023, by = "state_name") %>%
  mutate(
    acc_per_100k = 100000 * n_acc / pop
  ) %>% 
  ungroup()
```

### Total Accidents per State

```{r, echo=FALSE}
# --- bar chart showing accidents per state ---
acc_state_bar <- ggplot(
  data = acc_state,
  aes(
    x = reorder(State, -acc_per_100k),
    y = acc_per_100k
  )
) +
  geom_col(position = "dodge") +
  coord_flip() +
  labs(
    title = "Total Accidents Per 100K by State (2016 – 2023)",
    x = NULL,
    y = "Accidents Per 100K"
  )

# save figure to png for use in final report
f <- here("Rmd", "ieee", "figures", "acc_state_bar.png")
ggsave(f, acc_state_bar +
         theme_minimal(base_size = 7) +
         labs(title = NULL),
       width = 3.5, height = 4.5, units = "in")
knitr::plot_crop(f)

# --- choropleth map showing accidents per state ---
n_acc_state_map <- ggChoropleth(
  data = acc_state,
  aes(fill = acc_per_100k, map_id = state_name),
  map = states_map,
  interactive = T,
  title = "Total Accidents Per 100K Residents",
)

n_acc_state_map_png <- ggChoropleth(
  data = acc_state,
  aes(fill = acc_per_100k, map_id = state_name),
  map = states_map,
  interactive = FALSE,
)

# save figure to png for use in final report
f <- here("Rmd", "ieee", "figures", "n_acc_state_map.png")
ggsave(f, n_acc_state_map_png +
         scale_fill_gradient2(
           low = "white",
           mid = "grey",
           high = "red",
           midpoint = mean(acc_state$acc_per_100k),
           name = "Accidents/100K"
         ) +
         theme_void(base_size = 9) +
         theme(
           plot.margin = margin(0, 0, 0, 0, "pt"),
           legend.position = "bottom",
         ),
       width = 3.5, height = 2.5, units = "in")
knitr::plot_crop(f)

tbl_worst5_acc_state <- knitr::kable(
  head(acc_state %>%
         arrange(desc(acc_per_100k)) %>%
         select(state_name, acc_per_100k),
       5),
  col.names = c("State", "Accidents Per 100K"),
  caption = "Top 5 States with Most Accidents",
  format = "html")


f <- here("Rmd", "ieee", "figures", "worst-5-states-n-acc.png")
kableExtra::save_kable(tbl_worst5_acc_state %>% kable_styling(font_size = 24),f)
knitr::plot_crop(f)


# display figures in html
ggplotly(acc_state_bar, tooltip = c("x", "y"))
n_acc_state_map
tbl_worst5_acc_state
```

We can observe that, when adjusted for population, the following states: `r str_flatten_comma(head((acc_state %>% arrange(desc(acc_per_100k))), 5)$state_name)`, had the most accidents from 2016 to 2023.

### Average Severity by State

```{r, echo=FALSE}
mean_sev_state_bar <- ggplot(
  data = acc_state,
  aes(
    x = reorder(State, -mean_sev),
    y = mean_sev
  )
) +
  geom_col(position = "dodge") +
  coord_flip() +
  labs(
    title = "Average Accident Severity by State (2016 – 2023)",
    x = NULL,
    y = "Mean Severity (1 = least, 4 = most)"
  )

mean_sev_state_map <- ggChoropleth(
  data = acc_state,
  aes(fill = mean_sev, map_id = state_name),
  map = states_map,
  interactive = T,
  title = "Average Accident Severity by State (2016 – 2023)"
)

mean_sev_state_map_png <- ggChoropleth(
  data = acc_state,
  aes(fill = mean_sev, map_id = state_name),
  map = states_map,
  interactive = FALSE
)

f <- here("Rmd", "ieee", "figures", "mean-sev-state-map.png")
ggsave(f, mean_sev_state_map_png +
         scale_fill_gradient2(
           low = "white",
           mid = "grey",
           high = "red",
           midpoint = mean(acc_state$mean_sev),
           name = "Average Severity"
         ) +
         theme_void(base_size = 9) +
         theme(
           plot.margin = margin(0, 0, 0, 0, "pt"),
           legend.position = "bottom",
         ),
       width = 3.5, height = 2.5, units = "in")
knitr::plot_crop(f)

tbl_worst5_avg_sev <- knitr::kable(
  head(acc_state %>%
         arrange(desc(mean_sev)) %>%
         select(state_name, mean_sev),
         5),
  col.names = c("State", "Average Accident Severity"),
  caption = "Top 5 States with Worst Average Severity",
  format = "html"
)

f <- here("Rmd", "ieee", "figures", "mean-sev-state-bar.png")
ggsave(f, mean_sev_state_bar +
  theme_minimal(base_size = 7) ,
         width = 3.5, height = 4.5, units = "in")
knitr::plot_crop(f)

ggplotly(mean_sev_state_bar + theme_minimal(), tooltip = c("x", "y"))
mean_sev_state_map
tbl_worst5_avg_sev
```

While South Carolina had the most accidents per capita, the average severity was one of the lowest of all the states. The states that had the worst average severity were Georgia, Wisconsin, Rhode Island, Kentucky, and Colorado. While some states had a higher average severity than others, the largest difference in average severity was only `r round(max(acc_state$mean_sev)-min(acc_state$mean_sev), 2)`.

## Environmental Trends

```{r, echo=FALSE}
mean_temp_state_map <- ggChoropleth(
  data = acc_state,
  aes(fill = mean_temp, map_id = state_name),
  map = states_map,
  interactive = T
)

mean_temp_state_map_png <- ggChoropleth(
  data = acc_state,
  aes(fill = mean_temp, map_id = state_name),
  map = states_map,
  interactive = FALSE
)

f <- here("Rmd", "ieee", "figures", "mean-temp-state-map.png")
ggsave(f, mean_temp_state_map_png +
         scale_fill_gradient2(
           low = "white",
           mid = "grey",
           high = "red",
           midpoint = mean(acc_state$mean_temp),
           name = "Average Temperature (F)"
         ) +
         theme_void(base_size = 9) +
         theme(
           plot.margin = margin(0, 0, 0, 0, "pt"),
           legend.position = "bottom",
         ),
       width = 3.5, height = 2.5, units = "in")
knitr::plot_crop(f)

mean_temp_state_map
```

When we visualize the average accident temperature by state, we can observe that generally, accidents in northern states occur more frequently in cooler temperatures, while accidents in southern states occur more frequently in warmer temperatures.

```{r, echo=FALSE}
# we need to group by city because we can't assume that the weather will be the
#    same across the entire state
acc_state_day <- acc %>%
  group_by(State, City, date_) %>%
  summarize(
    n_acc = n(),
    mean_sev = mean(Severity),
    mean_precip = mean(`Precipitation(in)`),
    mean_temp = mean(`Temperature(F)`),
    precip_day = mean_precip > 0,
    hot_day = mean_temp > 80,
    cold_day = mean_temp < 50,
    .groups = "drop"
  )

state_cor <- acc_state_day %>%
  group_by(State) %>%
  summarise(
    r = cor(mean_temp, n_acc, use = "pair"),
    .groups = "drop"
  ) %>%
  mutate(state_name = state.name[match(State, state.abb)],
         map_id = state_name) %>%
  filter(!is.na(r))

state_cor_map <- ggChoropleth(
  data = state_cor,
  aes(fill = r, map_id = state_name),
  map = states_map,
  interactive = TRUE,
  title = "Correlation Between Temperature and Number of Accidents by State"
)

state_cor_map_png <- ggChoropleth(
  data = state_cor,
  aes(fill = r, map_id = state_name),
  map = states_map,
  interactive = FALSE
)

f <- here("Rmd", "ieee", "figures", "state-cor-map.png")
ggsave(f, state_cor_map_png +
         scale_fill_gradient2(
           name = "Average Temperature (F)"
         ) +
         theme_void(base_size = 9) +
         theme(
           plot.margin = margin(0, 0, 0, 0, "pt"),
           legend.position = "bottom",
         ),
       width = 3.5, height = 2.5, units = "in")
knitr::plot_crop(f)
```

We can observe that for most states, there doesn't seem to be a correlation between average temperature and number of accidents, but there are a few outliers. There is a slight positive correlation for South Dakota and a slight negative correlation for Wyoming.

```{r, echo=FALSE}
acc_day <- acc_state_day %>%
  group_by(date_) %>%
  summarise(
    # https://en.wikipedia.org/wiki/Weighted_arithmetic_mean
    # https://www.rdocumentation.org/packages/stats/versions/3.6.2/topics/weighted.mean
    mean_sev = weighted.mean(mean_sev, w = n_acc),
    mean_temp = weighted.mean(mean_temp, w = n_acc),
    mean_precip = weighted.mean(mean_precip, w = n_acc),
    n_acc = sum(n_acc),
    .groups = "drop"
  )

# create 10 degree temp range "buckets" and group accidents by temp range
#    (-20, -10], (-10, 0], (0, 10], etc.
temp_ranges <- acc_day %>%
  mutate(temp_range = cut(mean_temp, breaks = seq(-20, 120, 10))) %>%
  group_by(temp_range) %>%
  summarise(
    mean_n = mean(n_acc),
    mean_sev = mean(mean_sev),
    .groups = "drop"
  ) %>%
  filter(!is.na(temp_range))

# we can then plot the average number of accidents per day in each range
p <- ggplot(temp_ranges, aes(temp_range, mean_n, group = 1)) +
  geom_line() +
  geom_point() +
  labs(x = "Temperature range (F)",
       y = "Accidents per day",
       title = "Effect of Temperature on Number of Accidents")

f <- here("Rmd", "ieee", "figures", "acc-by-temp.png")
ggsave(f, p +
  theme_minimal(base_size = 7) ,
         width = 3.5, height = 2.5, units = "in")
knitr::plot_crop(f)

ggplotly(p + theme_minimal())
knitr::kable(
  (temp_ranges %>% arrange(desc(mean_n)) %>% select(temp_range, mean_n)),
  col.names = c("Temperature Range",
                "Number of Accidents"))
```

We can observe that accidents tend to be less likely at each extreme. Very cold temperatures and very hot temperatures see the least number of accidents. The temperature range 40-50 sees slightly more accidents than average.

```{r, echo=FALSE}
# mean severity per day by temp
p <- ggplot(temp_ranges, aes(temp_range, mean_sev, group = 1)) +
  geom_line() +
  geom_point() +
  labs(x = "Temperature range (F)",
       y = "Mean severity / day",
       title = "Effect of Temperature on Accident Severity")

f <- here("Rmd", "ieee", "figures", "sev-by-temp.png")
ggsave(f, p +
  theme_minimal(base_size = 7) ,
         width = 3.5, height = 2.5, units = "in")
knitr::plot_crop(f)

ggplotly(p + theme_minimal())

knitr::kable(
  (temp_ranges %>% arrange(desc(mean_sev)) %>% select(temp_range, mean_sev)),
  col.names = c("Temperature Range",
                "Average Severity"))
```

While fewer accidents occur at the temperature extremes, we can observe that the accidents that do occur are of a higher average severity. Accidents that occur when the temperature is between 10 and 20 degrees tend to have the highest severity.

## Statistical Analysis

### Correlation Analysis of Key Quantitative Features

The heatmap shows the correlation between quantitative features such as temperature, wind chill, visibility, precipitation, and severity. Temperature and wind chill were nearly perfectly correlated ($r = 0.99$), as expected. However, severity had only weak correlations with all other variables, suggesting that accident severity is influenced by additional factors beyond those measured here.

```{r}
numeric_data <- acc %>%
  select(Severity, visibility, temperature, wind_chill, precipitation) %>%
  na.omit()

cor_matrix <- cor(numeric_data)
cor_melted <- melt(cor_matrix)

p <- ggplot(cor_melted, aes(Var1, Var2, fill = value)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", 
                       midpoint = 0, limit = c(-1,1), space = "Lab", 
                       name="Correlation") +
  geom_text(aes(label = round(value, 2)), color = "black", size = 4) +
  labs(title = "Correlation Heatmap of Numerical Features", x = "", y = "")

f <- here("Rmd", "ieee", "figures", "cor-numerical.png")
ggsave(f, p +
  theme_minimal(base_size = 7) ,
         width = 3.5, height = 2.5, units = "in")
knitr::plot_crop(f)

ggplotly(p + theme_minimal())

```

### ANOVA on Accident Severity by Weather Condition

A one-way ANOVA was conducted to examine whether accident severity differs by weather condition. The results showed a statistically significant effect of weather on accident severity, $F(4, 1,\!814,\!823) = 18,\!549$, $p < .001$, indicating that the average severity of accidents varies across different weather conditions.

```{r}
anova_data <- acc %>%
  filter(!is.na(weather) & !is.na(Severity)) %>%
  filter(weather %in% c("Clear", "Cloudy", "Rain", "Snow", "Fog"))

# Run ANOVA
anova_result <- aov(Severity ~ weather, data = anova_data)

# Output ANOVA table
summary(anova_result)
```

### T-Tests on Severity and Frequency for Holidays

#### Severity on Specific Holidays T-Test

A Welch two-sample t-test was conducted to compare accident severity on specific holidays versus other days. The results showed a statistically significant difference in severity scores, $t(93,\!469) = 2.50$, $p = .0125$. The average severity on non-holidays ($M = 2.212$) was slightly higher than on holidays ($M = 2.208$), with a 95% confidence interval for the difference in means ranging from 0.0009 to 0.0073.

#### Frequency on Specific Holidays T-Test

A Welch two-sample t-test was also conducted to examine differences in the average number of accidents per day on holidays versus non-holidays. The results were statistically significant, $t(43.04) = 3.27$, $p = .0021$. The mean number of accidents per day was higher on non-holidays ($M = 2,\!947$) compared to holidays ($M = 2,\!173$), with a 95% confidence interval for the difference in means ranging from 297 to 1,!250.

```{r}
library(lubridate)
library(dplyr)
library(ggplot2)

# fixed date holidays
custom_holidays <- c("01-01", # New Year's Day
                     "07-04", # Independence Day
                     "12-25") # Christmas

# floating holidays
get_floating_holidays <- function(years) {
  holidays <- c()
  
  for (y in years) {
    # Thanksgiving: 4th Thursday in November
    thanksgiving <- as.Date(cut(as.Date(paste0(y, "-11-01")) + weeks(3), "week")) + 4
    while (weekdays(thanksgiving) != "Thursday") {
      thanksgiving <- thanksgiving + 1
    }
    
    # Memorial Day: last Monday in May
    memorial_day <- as.Date(paste0(y, "-05-31"))
    while (weekdays(memorial_day) != "Monday") {
      memorial_day <- memorial_day - 1
    }
    
    # Labor Day: first Monday in September
    labor_day <- as.Date(paste0(y, "-09-01"))
    while (weekdays(labor_day) != "Monday") {
      labor_day <- labor_day + 1
    }
    
    holidays <- c(holidays, thanksgiving, memorial_day, labor_day)
  }
  
  as.Date(holidays)
}

# full holiday list
years <- 2016:2023
floating_days <- get_floating_holidays(years)

fixed_days <- do.call(c, lapply(years, function(y) {
  as.Date(paste0(y, "-", custom_holidays))
}))

# Combine all holidays
specific_holidays <- sort(c(fixed_days, floating_days))

# Flag holidays in accident data
acc$holiday_specific <- acc$date_ %in% specific_holidays


# T-Test: Severity
t_test_severity <- t.test(Severity ~ holiday_specific, data = acc)
print("T-test on Severity (Specific Holidays):")
print(t_test_severity)

# Frequency per day
acc_day <- acc %>%
  group_by(date_) %>%
  summarise(n_acc = n(), holiday_specific = any(holiday_specific)) %>%
  ungroup()

t_test_freq <- t.test(n_acc ~ holiday_specific, data = acc_day)
print("T-test on Frequency (Specific Holidays):")
print(t_test_freq)

```

#### Visualization: Severity on Holidays

Although the difference is small, the chart shows a **slightly higher average severity for accidents on non-holidays** compared to holidays. The mean severity was **2.212 on non-holidays** and **2.208 on holidays**. The corresponding Welch t-test ($t(93,\!469) = 2.50$, $p = .0125$) confirms that this difference is **statistically significant**, although not practically large. This suggests that while there are fewer accidents on holidays, they are not necessarily more or less severe.

```{r}
p <- acc %>%
  group_by(holiday_specific) %>%
  summarise(mean_severity = mean(Severity)) %>%
  ggplot(aes(x = holiday_specific, y = mean_severity, fill = holiday_specific)) +
  geom_col() +
  labs(title = "Average Severity on Specific Holidays vs. Other Days", y = "Avg Severity", x = "Is Specific Holiday")

f <- here("Rmd", "ieee", "figures", "sev-holiday.png")
ggsave(f, p +
  theme_minimal(base_size = 7) ,
         width = 3.5, height = 2.5, units = "in")
knitr::plot_crop(f)

ggplotly(p + theme_minimal())
```

#### Visualization: Frequency of Accidents on Holidays

The bar chart clearly shows that the **average number of accidents per day is significantly lower on specific holidays** compared to non-holiday dates. On average, there were around **2,173 accidents per day on holidays** versus **2,947 on non-holidays**. This visual supports the results of the Welch two-sample t-test ($t(43.04) = 3.27$, $p = .0021$), confirming that this difference is **statistically significant**. The lower volume on holidays may reflect reduced traffic due to time off from work and school.

```{r}
p <- acc_day %>%
  group_by(holiday_specific) %>%
  summarise(mean_accidents = mean(n_acc)) %>%
  ggplot(aes(x = holiday_specific, y = mean_accidents, fill = holiday_specific)) +
  geom_col() +
  labs(title = "Average Daily Accidents: Specific Holidays vs. Other Days", y = "Avg Accidents/Day", x = "Is Specific Holiday")

f <- here("Rmd", "ieee", "figures", "acc-holiday2.png")
ggsave(f, p +
  theme_minimal(base_size = 7) ,
         width = 3.5, height = 2.5, units = "in")
knitr::plot_crop(f)

ggplotly(p + theme_minimal())
```

```{r echo=TRUE, message=FALSE, warning=FALSE}
t.test(data = acc_state_day, mean_sev ~ precip_day)
t.test(data = acc_state_day, mean_sev ~ hot_day)
t.test(data = acc_state_day, mean_sev ~ cold_day)

t.test(data = acc_state_day, n_acc ~ precip_day)
t.test(data = acc_state_day, n_acc ~ hot_day)
t.test(data = acc_state_day, n_acc ~ cold_day)
```
